### 2020-02-13 CalNotes.tex
Update after second estimator and Bayesian

\documentclass[10pt]{article}

@@

Data-driven approaches to the derivation of effective stochastic coarse-grained dynamics has become a standard tool in different areas such as molecular dynamics (CITE) or atmosphere and ocean science \cite{KKP15,CoP09,YMV19}. When fitting data to a low dimensional coarse-grained model, one is faced with the problem of model misspecification; indeed, the data, coming from the full dynamics, is compatible with the coarse-grained model only at the time scales at which the effective dynamics is valid. This is related to the issue of high frequency data that is arises in econometrics \cite{AiJ14,AMZ05,ZMA05,OSP10}.

@@

For simple models in molecular dynamics, the effect of model misspecification was studied in a series of papers \cite{PaS07,PPS09,PPS12,GaS17,GaS18} under the assumption of scale separation. In particular, for Brownian particles moving in two-scale potentials it was shown that, when fitting data from the full dynamics to the homogenised equation, the maximum likelihood estimator (MLE) is asymptotically biased \cite[Theorem 3.4]{PaS07}. To be more precise, in the large sample size limit the MLE converges to the coefficients of the unhomogenised equation, rather than to those of the homogenised one. The bias of the MLE estimator can be eliminated by subsampling at an appropriate rate, which lies between the two characteristic time scales of the problem \cite[Theorems 3.5 and 3.6]{PaS07}. This is similar to the estimation of the integrated stochastic volatility in the presence of market micro-structure noise, where the data has to be subsampled at an appropriate rate \cite{AMZ05,ZMA05}. The correct subsampling rate can be in some instances rather extreme with respect to the frequency of the data and lead to get rid of more than $99\%$ of the data. As the intuition suggests, this increases significantly the bias of the estimator, which is usually taken care of with additional bias corrections and variance reduction procedures.

@@

It is natural to reinterpret the problem of inferring the drift parameter of multiscale diffusion processes under a Bayesian perspective \cite{PSV09,PSZ13}. In particular, the form of the likelihood function guarantees, under a Gaussian prior hypothesis, that the posterior distribution is itself a Gaussian. Therefore, one can simply compute the mean and covariance of the posterior distribution to obtain a full quantification of the inference process. The Bayesian approach has been adopted in several works concerning single and multiscale PDEs, too, and has proved itself to be versatile and efficient in this framework (see, e.g., \cite{AbD18, AGZ19}). Moreover, infinite-dimensional inverse problems are intrinsically ill-posed, and interpreting them in a Bayesian manner is an efficient tool for regularizing them and guaranteeing their well-posedness \cite{Stu10, DaS16}. The regularization property of a Bayesian approach has been recently demonstrated for inverse problems involving multiscale PDE \cite{AbD18}. In this work, we investigate the effects of inserting our filtering methodology in a Bayesian framework. In particular, we analyse the effects of modifying the likelihood function inserting the filtered trajectory, showing how this corrects the faulty asymptotic behaviour which would be present in case no pre-processing would be applied.

\subsection{Outline}

@@

Let $\epl > 0$ and let us consider the one-dimensional multiscale stochastic differential equation (SDE)
\begin{equation}\label{eq:SDE_MS}
	\d X_t^\epl = -\alpha \cdot V'(X_t^\epl) \dd t - \frac1\epl p'\left(\frac{X_t^\epl}\epl\right) \dd t + \sqrt{2\sigma} \dd W_t,
\end{equation}
where $\alpha_i \in \R$, $i = 1, \ldots, N$ and $\sigma > 0$ are the drift and diffusion coefficients respectively and $W_t$ is a standard one-dimensional Brownian motion. The functions $V\colon \R \to \R^N$ and $p\colon \R \to \R$ are slow and fast potentials driving the dynamics of the solution $X_t^\epl$. Given $N \in \N_{>0}$, we assume the slow potential to be of the form
\begin{equation}\label{eq:Potential}
	V(x) = \begin{pmatrix} V_1(x) & V_2(x) & \cdots & V_N(x) \end{pmatrix}^\top,
\end{equation}
for smooth functions $V_i\colon \R \to \R$, $i = 1, \ldots, N$. Moreover, we assume $p$ to be smooth and periodic of period $L$. Theory of homogenization \cite{BLP78} guarantees the existence of an SDE of the form
\begin{equation}\label{eq:SDE_HOM}
	\d X_t^0 = - A \cdot V'(X_t^0) \dd t + \sqrt{2\Sigma} \dd W_t,
\end{equation}
where $W_t$ is the same Brownian motion and where the fast dynamics have been eliminated, such that $X_t^\epl \to X_t^0$ for $\epl\to 0$ in law as random variables in $\mathcal C^0((0, T), \R)$. In particular, we have $A_i = K\alpha_i$ and $\Sigma = K \sigma$, where the coefficient $K$ is given by the formula
\begin{equation}\label{eq:K_HOM}
	K = \int_0^L (1 + \Phi'(y))^2 \, \mu(\d y),
\end{equation}
with 
\begin{equation}
	\mu(\d y) = Z^{-1} e^{-p(y)/\sigma} \dd y, \quad\text{where}\quad Z = \int_0^L e^{-p(y)/\sigma} \dd y,
\end{equation}
and where the function $\Phi$ is the solution of the elliptic partial differential equation
\begin{equation}
	-p'(y)\Phi'(y) + \sigma \Phi''(y) = p'(y), \quad 0 \leq y \leq L,
\end{equation}
endowed with periodic boundary conditions.

Let $T > 0$ and let $X^\epl \defeq (X^\epl_t, 0\leq t, \leq T)$ be a realization of the solution of \eqref{eq:SDE_MS} up to final time. In order to estimate the drift coefficient $A$ of the homogenized equation we maximize the likelihood function
\begin{equation}
p(X^\epl \mid A) = \exp\left(-\frac{I(X^\epl\mid A)}{2\Sigma} \right), 
\end{equation}
where Girsanov formula allows to write explicitly the log-likelihood as
\begin{equation}
I(X^\epl \mid A) = \int_0^T A \cdot V'(X^\epl_t) \dd X^\epl_t + \frac12 \int_0^T \left( A \cdot V'(X^\epl_t) \right)^2 \dd t.
\end{equation}
Therefore, the maximum likelihood estimator can be formally computed in closed form as
\begin{equation}
	\widehat A^\epl(T) \defeq \arg \min_{A \in \R^N} I(X^\epl \mid A) = M^{-1}h,
\end{equation}
where $M\in\R^{N\times N}$ and $h\in\R^N$ are defined as
\begin{equation}
M = \frac1{2\Sigma T} \int_0^T V'(X_t) \otimes V'(X_t) \dd t, \quad h = \frac1{2\Sigma T} \int_0^T V'(X_t) \dd X_t,
\end{equation}
and where $\otimes$ denotes the tensor product in $\R^N$.
Let us now state the assumptions which will be employed throughout the rest of our work. In particular, we consider the dissipative setting of \cite{PaS07}.
\begin{assumption}\label{as:regularity} The potentials $p$ and $V$ satisfy
	\begin{enumerate}
		\item $p \in \mathcal C^\infty(\R, \R) \cap L^\infty(\R, \R)$,
		\item\label{as:regularity_diss} the potential $V \in \mathcal C^\infty(\R, \R)$ and there exist $a,b > 0$ such that
		\begin{equation}
		-V'(x)x \leq a - bx^2.
		\end{equation} 
		Moreover, for all $T > 0$, the symmetric matrix $M$ is positive definite and its minimum eigenvalue satisfies $\lambda_{\min}(M) \geq \bar \lambda > 0$.
	\end{enumerate}
\end{assumption}
Under these assumptions, the estimator $\widehat A^\epl(T)$ is indeed well-posed. Moreover, these assumptions guarantee that the process $X^\epl$ is geometrically ergodic, and

@@

In the previous section, we have shown that posterior distributions over the drift coefficient of the homogenized equation are asymptotically incorrect if multiscale data are replaced into the expression of the likelihood. Hence, the need of pre-processing the data is highlighted. In particular, we consider a filtering approach. Let $k \colon \R^+ \times \R^+ \to \R$ be a kernel function and consider the process $Z^\epl \defeq (Z^{\epl,k}_t, 0 \leq t \leq T)$ defined by the weighted average
\begin{equation}\label{eq:ZDef}
	Z^{\epl}_t \defeq \int_0^t k(t, s)X^\epl_s \dd s.
\end{equation}
Let $\beta, \delta > 0$ and let us consider a family of exponential kernel functions defined as
\begin{equation}\label{eq:filter}
	k(t, s) = C_\beta \delta^{-1/\beta} e^{-(t-s)^\beta/\delta},
\end{equation}
where $C_{\beta}$ is a normalizing constant given by
\begin{equation}
	C_\beta = \beta \, \Gamma(1/\beta)^{-1},
\end{equation}
and where $\Gamma(\cdot)$ is the gamma function. In the following, only in case $\beta = 1$ a rigorous analysis is carried on. Nonetheless, numerical experiments show that for higher values of $\beta$ the performances of estimators computed employing the filter are more robust and qualitatively better.

Given a trajectory $X^\epl \defeq (X^\epl_t, 0\leq t \leq T)$, it is relatively inexpensive to compute $Z^\epl$ from a computational standpoint. In particular, the process $Z^\epl$ is the convolution of the kernel with the process $X^\epl$. Hence, computational tools based on the Fast Fourier Transform (FFT) exist and allow to compute $Z^\epl$ fast component-wise.

@@

In the following, we consider the filter with $\beta = 1$, i.e.,
\begin{equation}\label{eq:filter_beta1}
	k(t,s) = \frac{1}{\delta} e^{-\frac{t-s}{\delta}},
\end{equation}
since in this case it holds
\begin{equation}
	\d Z^\epl_t = k(t,t) X^\epl_t \dd t + \int_0^t \partial_t k(t,s) X^\epl_s \dd s \dd t = \frac{1}{\delta} \left ( X^\epl_t - Z^\epl_t \right ) \dd t.
\end{equation}
Considering the processes $X^\epl$ and $Z^\epl$ together, we obtain a system of stochastic differential equations
\begin{equation}
\label{eq:systemSDE}
\begin{aligned}
\d X_t^\epl &= -\alpha \cdot V'(X_t^\epl) \dd t - \frac1\epl p'\left(\frac{X_t^\epl}\epl\right) + \sqrt{2\sigma} \dd W_t, \\
\d Z^\epl_t &= \frac{1}{\delta} \left ( X^\epl_t - Z^\epl_t \right ) \dd t.
\end{aligned}
\end{equation}
The following result guarantees that the measure induced by the joint process $(X_t^\epl, Z_t^\epl)^\top$ admits a smooth density.

\begin{lemma}\label{lem:density} Let $(X_t^\epl, Z_t^\epl)^\top$ be the solution of \eqref{eq:systemSDE}. Then the measure $\mu^\epl$ of the process admits a smooth density $\rho^\epl$ with respect to the Lebesgue measure.
%	 satisfying
%	\begin{equation}
%	\begin{aligned}
%	\partial_t\rho^\epl(t,x,z) -\sigma \partial^2_{xx} \rho^\epl(t,x,z) &- \partial_x \left(\left ( \alpha \cdot V'(x) + \frac{1}\epl p' \left ( \frac{x}\epl \right ) \right )\rho^\epl(t,x,z)\right) \\
%	&- \frac{1}{\delta}\partial_z\left((z - x) \rho^\epl(t,x,z)\right) = 0.
%	\end{aligned}
%	\end{equation} 
\end{lemma}

\begin{proof} We have to show that the joint process solution to \eqref{eq:systemSDE} is hypo-elliptic. Let us define the function $f\colon \R \to \R$ as 
	\begin{equation}
		f(x) = -\alpha \cdot V'(x) - \frac1\epl p'\left(\frac{x}\epl\right).
	\end{equation}
	Then, the generator of the process $(X_t^\epl, Z_t^\epl)^\top$ is given by
	\begin{equation}
		\mathcal L = f \partial_x + \sigma \partial_{xx}^2 + \frac1\delta (x - z)\partial_z \eqdef \mathcal X_0 + \sigma \mathcal X_1^2, 
	\end{equation}
	where 
	\begin{equation}
		\mathcal X_0 = f \partial_x + \frac1\delta (x - z)\partial_z, \quad \mathcal X_1 = \partial_x.
	\end{equation}
	The commutator $[\mathcal X_0, \mathcal X_1]$ applied to a test function $v$ then gives
	\begin{equation}
	\begin{aligned}
		[\mathcal X_0, \mathcal X_1]v &= f \partial_x^2 v + \frac1\delta (x - z) \partial_x \partial_z v  - \partial_x\left(f\partial_x v + \frac1\delta (x - z)\partial_z v\right)\\
		&= -\partial_x f \partial_x v - \frac1\delta \partial_z v.
	\end{aligned}
	\end{equation}
	Consequently, 
	\begin{equation}
		\mathrm{Lie}\left(\mathcal X_1, [\mathcal X_0, \mathcal X_1]\right) = \mathrm{Lie} \left(\partial_x, -\partial_x f \partial_x - \frac1\delta\partial_z\right),
	\end{equation}
	which spans $T_x\R^2$. The desired result then follows from Hörmander's theorem (see e.g. \cite[Chapter 6]{Pav14}).
\end{proof}

Having established that the law of the joint process admits a smooth density with respect to the Lebesgue measure, we now show in the following Lemma a property of ergodicity.

\begin{lemma}\label{lem:ergodicity} Let Assumption \ref{as:regularity} hold and let $b > 0$ be given in Assumption \ref{as:regularity}\ref{as:regularity_diss}. Then, if $\delta > 1/(4b)$, the process $(X_t^\epl, Z_t^\epl)^\top$ solution of \eqref{eq:systemSDE} is geometrically ergodic, i.e., there exists $C, \lambda > 0$ such that for all measurable $f$ such that for some integer $p > 0$ 
	\begin{equation}
		f(x) \leq 1 + \abs{x}^p,
	\end{equation}
	it holds
	\begin{equation}
		\abs{\E f(X^\epl_t, Z^\epl_t) - \int_\R\int_\R f(x, z) \rho^\epl(x, z) \dd x \dd z} \leq C\left(1 + \norm{\begin{pmatrix} X^\epl_0 \\ Z^\epl_0 \end{pmatrix}}^p \right)e^{-\lambda t},
	\end{equation}
	where $\rho^\epl$ is the solution to the stationary Fokker--Planck equation
	\begin{equation}
	\label{eq:FPsystem}
	\sigma \partial^2_{xx} \rho^\epl(x,z) +  \partial_x \left( \left( \alpha \cdot V'(x) + \frac{1}{\epl} p' \left ( \frac{x}\epl \right ) \right ) \rho^\epl(x,z) \right) + \frac{1}{\delta} \partial_z \left((z - x) \rho^\epl(x,z)\right) = 0.
	\end{equation}
\end{lemma}

>>>>

\documentclass[10pt]{article}
%\documentclass{siamart1116}

@@

Data-driven approaches to the derivation of effective stochastic coarse-grained dynamics has become a standard tool in different areas such as molecular dynamics or atmosphere and ocean science \cite{KKP15,CoP09,YMV19}. When fitting data to a low dimensional coarse-grained model, one is faced with the problem of model misspecification; indeed, the data, coming from the full dynamics, is compatible with the coarse-grained model only at the time scales at which the effective dynamics is valid. This is related to the issue of high-frequency data that arises in econometrics \cite{AiJ14,AMZ05,ZMA05,OSP10}.

@@

For simple models in molecular dynamics, the effect of model misspecification was studied in a series of papers \cite{PaS07,PPS09,PPS12,GaS17,GaS18} under the assumption of scale separation. In particular, for Brownian particles moving in two-scale potentials it was shown that, when fitting data from the full dynamics to the homogenised equation, the maximum likelihood estimator (MLE) is asymptotically biased \cite[Theorem 3.4]{PaS07}. To be more precise, in the large sample size limit the MLE converges to the coefficients of the unhomogenised equation, rather than to those of the homogenised one. The bias of the MLE can be eliminated by subsampling at an appropriate rate, which lies between the two characteristic time scales of the problem \cite[Theorems 3.5 and 3.6]{PaS07}. This is similar to the estimation of the integrated stochastic volatility in the presence of market micro-structure noise, where the data has to be subsampled at an appropriate rate \cite{AMZ05,ZMA05}. The correct subsampling rate can be in some instances rather extreme with respect to the frequency of the data and lead to get rid of more than $99\%$ of the data. As the intuition suggests, this increases significantly the bias of the estimator, which is usually taken care of with additional bias corrections and variance reduction procedures.

@@

It is natural to reinterpret the problem of inferring the drift parameter of multiscale diffusion processes under a Bayesian perspective \cite{PSV09,PSZ13}. In particular, the form of the likelihood function guarantees, under a Gaussian prior hypothesis, that the posterior distribution is itself a Gaussian. Therefore, one can simply compute the mean and covariance of the posterior distribution to obtain a full uncertainty quantification of the inference process. The Bayesian approach has been adopted in several works concerning single and multiscale PDEs, too, and has proved itself to be versatile and efficient in this framework (see, e.g., \cite{AbD18, AGZ19}). Moreover, infinite-dimensional inverse problems are intrinsically ill-posed, and interpreting them in a Bayesian manner is an efficient tool for regularizing them and guaranteeing their well-posedness \cite{Stu10, DaS16}. The regularization property of a Bayesian approach has been recently demonstrated for inverse problems involving multiscale PDEs \cite{AbD18}. In this work, we investigate the effects of inserting our filtering methodology in a Bayesian framework. In particular, we analyse the effects of modifying the likelihood function inserting the filtered trajectory, showing how this corrects the faulty asymptotic behaviour which would be present in case no pre-processing would be applied.

@@

In this section, we introduce the class of diffusion processes which we treat in this paper and the classical methodology employed for the estimation of the drift. Let $\epl > 0$ and let us consider the one-dimensional multiscale stochastic differential equation (SDE)
\begin{equation}\label{eq:SDE_MS}
	\d X_t^\epl = -\alpha \cdot V'(X_t^\epl) \dd t - \frac1\epl p'\left(\frac{X_t^\epl}\epl\right) \dd t + \sqrt{2\sigma} \dd W_t,
\end{equation}
where, given a positive integer $N$, we have that $\alpha \in \R^N$ and $\sigma > 0$ are the drift and diffusion coefficients respectively and $W_t$ is a standard one-dimensional Brownian motion. The functions $V\colon \R \to \R^N$ and $p\colon \R \to \R$ correspond to the fast-scale and the slow-scale confining potentials. In particular, we assume 
\begin{equation}\label{eq:Potential}
	V(x) = \begin{pmatrix} V_1(x) & V_2(x) & \cdots & V_N(x) \end{pmatrix}^\top,
\end{equation}
for smooth functions $V_i\colon \R \to \R$, $i = 1, \ldots, N$. Moreover, we assume $p$ to be smooth and periodic of period $L$. Theory of homogenization \cite[Chapter 3]{BLP78} guarantees the existence of an SDE of the form
\begin{equation}\label{eq:SDE_HOM}
	\d X_t = - A \cdot V'(X_t) \dd t + \sqrt{2\Sigma} \dd W_t,
\end{equation}
where $W_t$ is the same Brownian motion as in \eqref{eq:SDE_MS}, such that $X_t^\epl \to X_t$ for $\epl\to 0$ in law as random variables in $\mathcal C^0((0, T); \R)$. In particular, we have $A_i = K\alpha_i$ and $\Sigma = K \sigma$, where the coefficient $K$ is given by the formula
\begin{equation}\label{eq:K_HOM}
	K = \int_0^L (1 + \Phi'(y))^2 \, \mu(\d y),
\end{equation}
with 
\begin{equation}
	\mu(\d y) = \frac1Z e^{-p(y)/\sigma} \dd y, \quad\text{where}\quad Z = \int_0^L e^{-p(y)/\sigma} \dd y,
\end{equation}
and where the function $\Phi$ is the solution of the elliptic partial differential equation
\begin{equation}
	-p'(y)\Phi'(y) + \sigma \Phi''(y) = p'(y), \quad 0 \leq y \leq L,
\end{equation}
endowed with periodic boundary conditions.

We now briefly present the classical methodology for estimating the drift coefficient. Let $T > 0$ and let $X^\epl \defeq (X^\epl_t, 0\leq t, \leq T)$ be a realization of the solution of \eqref{eq:SDE_MS} up to final time. Girsanov's change of measure formula applied to \eqref{eq:SDE_HOM} allows to write the likelihood of $X^\epl$ given a drift coefficient $A$ of \eqref{eq:SDE_HOM} as
\begin{equation}\label{eq:Likelihood}
p(X^\epl \mid A) = \exp\left(-\frac{I(X^\epl\mid A)}{2\Sigma} \right), 
\end{equation}
where 
\begin{equation}
I(X^\epl \mid A) = \int_0^T A \cdot V'(X^\epl_t) \dd X^\epl_t + \frac12 \int_0^T \left( A \cdot V'(X^\epl_t) \right)^2 \dd t.
\end{equation}
Minimizing the functional $I(X^\epl \mid A)$ with respect to $A$ therefore gives the maximum likelihood estimator (MLE) of $A$, which can be formally computed in closed form as
\begin{equation}\label{eq:MLE}
	\widehat A^\epl(T) \defeq \arg \min_{A \in \R^N} I(X^\epl \mid A) = M^{-1}h,
\end{equation}
where $M\in\R^{N\times N}$ and $h\in\R^N$ are defined as
\begin{equation}\label{eq:MandH}
M = \frac1{2\Sigma T} \int_0^T V'(X^\epl_t) \otimes V'(X^\epl_t) \dd t, \quad h = \frac1{2\Sigma T} \int_0^T V'(X^\epl_t) \dd X^\epl_t,
\end{equation}
and where $\otimes$ denotes the tensor product in $\R^N$.
Let us now state the assumptions which will be employed throughout the rest of our work. In particular, we consider the same dissipative setting as \cite[Assumption 3.1]{PaS07}.
\begin{assumption}\label{as:regularity} The potentials $p$ and $V$ satisfy
	\begin{enumerate}
		\item $p \in \mathcal C^\infty(\R) \cap L^\infty(\R)$ and is $L$-periodic for some $L > 0$,
		\item\label{as:regularity_diss} $V_i \in \mathcal C^\infty(\R)$ for all $i=1, \ldots, N$ is polynomially bounded from above and bounded from below, and there exist $a,b > 0$ such that
		\begin{equation}
		-\sum_{i=1}^N V_i'(x)x \leq a - bx^2.
		\end{equation} 
		\item for all $T > 0$, the symmetric matrix $M$ is positive definite and there exists $\bar \lambda > 0$ such that $\lambda_{\min}(M) \geq \bar \lambda$.
	\end{enumerate}
\end{assumption}

Under these assumptions, the MLE given in \eqref{eq:MLE} is indeed the unique minimizer of the likelihood function, as shown in \cite[Theorem 2.4]{PSV09}.

Given the convergence of $X^\epl_t \to X_t$ in the space of continuous stochastic processes, one would expect that the MLE \eqref{eq:MLE} would be asymptotically unbiased for the drift coefficient $A$ of the homogenized equation \eqref{eq:SDE_HOM}. Instead, it is possible to prove that in the asymptotic limit for $T \to \infty$ and $\epl \to 0$, the MLE tends to the drift coefficient $\alpha$ of the unhomogenized equation \eqref{eq:SDE_MS}. We report here this result, whose proof can be found for the case $N = 1$ in \cite[Theorem 3.4]{PaS07}. Let us remark that the proof for $N > 1$ follows directly from the one-dimensional case.

\begin{theorem}\label{thm:Bias} Let Assumption \ref{as:regularity} hold and let $X^\epl_0$ be distributed according to the invariant measure of the process $X_t^\epl$ solution of \eqref{eq:SDE_MS}. Then
	\begin{equation}
		\lim_{\epl \to 0}\lim_{T \to \infty} \widehat A^\epl(T) = \alpha, \quad \text{a.s.},
	\end{equation}
	where $\alpha$ is the drift coefficient of equation \eqref{eq:SDE_MS}.
\end{theorem}

As anticipated in the introduction, the main tool for obtaining unbiased estimators in the literature is subsampling the data. In particular, let the dimension of the parameter $N = 1$, let $\delta > 0$ and let $T = n\delta$ with $n$ a positive integer. Then, a subsampled estimator for $A$ is given by
\begin{equation}
	\widehat A^\epl_\delta(T) = - \frac{\sum_{j=0}^{n-1} V'(X^\epl_{j\delta})\left(X^\epl_{(j+1)\delta} - X^\epl_{j\delta}\right)}{\delta \sum_{j=0}^{n-1} V'(X^\epl_{j\delta})^2},
\end{equation}
which is a discretized version of $\widehat A^\epl(T)$. It is possible to show \cite[Theorem 3.5]{PaS07} that choosing $\delta = \epl^\zeta$ with $\zeta \in (0, 1)$ and if $T$ is sufficiently big, then $\widehat A^\epl_\delta(T)$ is an asymptotically unbiased estimator of $A$ with respect to $T \to \infty$ and $\epl \to 0$, in probability. Despite being widely employed in practice, estimators based on subsampling present some drawbacks, as discussed in the introduction. In the following, we will introduce and analyse a novel approach for the drift estimation.

\begin{remark} Let us remark that for enhancing the clarity of the exposition, in this article we chose to focus on the case of a multi-dimensional parameter in the setting of one-dimensional diffusion processes. In fact, all the theory we present in the following could be generalized to the case of $d$-dimensional SDEs equivalent to \eqref{eq:SDE_MS}, which can be written as
	\begin{equation}
		\d X_t^\epl = - \sum_{i=1}^N \alpha_i \nabla V_i(X_t^\epl) \dd t - \frac1\epl \nabla p\left(\frac{X_t^\epl}{\epl}\right) \dd t + \sqrt{2\sigma} \dd W_t,
	\end{equation}
	where $W_t$ is a standard $d$-dimensional Brownian motion. The proof of all results below should be slightly modified, but we verified that all arguments still hold true in the $d$-dimensional case. 
\end{remark}

@@

In this section, we introduce and analyse a novel filtering approach to solve the biasedness issue highlighted by Theorem \ref{thm:Bias}. Let $\beta, \delta > 0$ and let us consider a family of exponential kernel functions $k \colon \R^+ \to \R$ defined as
\begin{equation}\label{eq:filter}
k(r) = C_\beta \delta^{-1/\beta} e^{-r^\beta/\delta},
\end{equation}
where $C_{\beta}$ is a normalizing constant given by
\begin{equation}
C_\beta = \beta \, \Gamma(1/\beta)^{-1},
\end{equation}
and where $\Gamma(\cdot)$ is the gamma function. We consider the process $Z^\epl \defeq (Z^\epl_t, 0 \leq t \leq T)$ defined by the weighted average
\begin{equation}\label{eq:ZDef}
	Z^{\epl}_t \defeq \int_0^t k(t - s)X^\epl_s \dd s.
\end{equation}
The process $Z^\epl$ can be interpreted as a smoothed version of the original trajectory $X^\epl$. In fact, in the field of signal processing the kernel \eqref{eq:filter} belongs to the class of low-pass linear time-invariant filters, which cut the high frequencies in a signal to highlight its slowest components. In the following, only in case $\beta = 1$ a rigorous analysis is carried on. Nonetheless, numerical experiments show that for higher values of $\beta$ the performances of estimators computed employing the filter are more robust and qualitatively better.

\begin{remark} Given a trajectory $X^\epl \defeq (X^\epl_t, 0\leq t \leq T)$, it is relatively inexpensive to compute $Z^\epl$ from a computational standpoint. In particular, the process $Z^\epl$ is the truncated convolution of the kernel with the process $X^\epl$. Hence, computational tools based on the Fast Fourier Transform (FFT) exist and allow to compute $Z^\epl$ fast component-wise. Moreover, the process $Z_t^\epl$ can be computed, in case $\beta = 1$, in a recursive manner and therefore ``online''.
\end{remark}

In the rest of this section we will focus on the properties of the process $Z^\epl$ when it is considered together with the original process $X^\epl$. In particular, we focus on ergodic properties and multiscale convergence. Finally, we present and analyse an unbiased estimator obtained with the filtering process.

@@

Let us consider the filtering kernel \eqref{eq:filter} with $\beta = 1$, i.e.,
\begin{equation}\label{eq:filter_beta1}
	k(r) = \frac{1}{\delta} e^{-r/\delta}.
\end{equation}
In this case, Leibniz integral rule yields the equality
\begin{equation}
	\d Z^\epl_t = k(0) X^\epl_t \dd t + \int_0^t k'(t-s) X^\epl_s \dd s \dd t = \frac{1}{\delta} \left ( X^\epl_t - Z^\epl_t \right ) \dd t,
\end{equation}
which can be interpreted as an ordinary differential equation for $Z_t^\epl$ driven by the stochastic signal $X^\epl$. Considering the processes $X^\epl$ and $Z^\epl$ together, we obtain the system of two one-dimensional SDEs
\begin{equation}
\label{eq:systemSDE}
\begin{aligned}
\d X_t^\epl &= -\alpha \cdot V'(X_t^\epl) \dd t - \frac1\epl p'\left(\frac{X_t^\epl}\epl\right) + \sqrt{2\sigma} \dd W_t, \\
\d Z^\epl_t &= \frac{1}{\delta} \left ( X^\epl_t - Z^\epl_t \right ) \dd t.
\end{aligned}
\end{equation}
The first ingredient for verifying the ergodic properties of the two-dimensional process $(X^\epl, Z^\epl)^\top \defeq ((X^\epl_t, Z^\epl_t)^\top, 0 \leq t \leq T)$ is verifying that the measure induced by the stochastic process admits a smooth density with respect to the Lebesgue measure. Since noise is present only on the first component, this is a consequence of the theory of hypo-ellipticity, as summarized in the following Lemma.

\begin{lemma}\label{lem:density} Let $(X^\epl, Z^\epl)^\top$ be the solution of \eqref{eq:systemSDE} and let $\mu^\epl_t$ be the measure induced by the couple at time $t$. Then, the measure $\mu^\epl_t$ admits a smooth density $\rho^\epl_t$ with respect to the Lebesgue measure.
\end{lemma}

\begin{proof} We have to show that the joint process solution to \eqref{eq:systemSDE} is hypo-elliptic. Denoting as $f\colon \R \to \R$ the function 
	\begin{equation}
		f(x) = -\alpha \cdot V'(x) - \frac1\epl p'\left(\frac{x}\epl\right),
	\end{equation}
	the generator of the process $(X^\epl, Z^\epl)^\top$ is given by
	\begin{equation}
		\mathcal L = f \partial_x + \sigma \partial_{xx}^2 + \frac1\delta (x - z)\partial_z \eqdef \mathcal X_0 + \sigma \mathcal X_1^2, 
	\end{equation}
	where 
	\begin{equation}
		\mathcal X_0 = f \partial_x + \frac1\delta (x - z)\partial_z, \quad \mathcal X_1 = \partial_x.
	\end{equation}
	The commutator $[\mathcal X_0, \mathcal X_1]$ applied to a test function $v$ then gives
	\begin{equation}
	\begin{aligned}
		[\mathcal X_0, \mathcal X_1]v &= f \partial_x^2 v + \frac1\delta (x - z) \partial_x \partial_z v  - \partial_x\left(f\partial_x v + \frac1\delta (x - z)\partial_z v\right)\\
		&= -\partial_x f \partial_x v - \frac1\delta \partial_z v.
	\end{aligned}
	\end{equation}
	Consequently, 
	\begin{equation}
		\mathrm{Lie}\left(\mathcal X_1, [\mathcal X_0, \mathcal X_1]\right) = \mathrm{Lie} \left(\partial_x, -\partial_x f \partial_x - \frac1\delta\partial_z\right),
	\end{equation}
	which spans $T_x\R^2$. The desired result then follows from Hörmander's theorem (see e.g. \cite[Chapter 6]{Pav14}).
\end{proof}

Once it is established that the law of the process admits a smooth density for all times $t>0$, which satisfies a time-dependent Fokker--Planck equation, we are interested in the limiting properties of this law. In particular, we know that the process $X^\epl$ alone is geometrically ergodic \cite[Theorem 4.4]{MSH02}, and we wish the couple $(X^\epl, Z^\epl)^\top$ to be endowed with the same property. The following Lemma guarantees that the couple is indeed geometrically ergodic exploiting an argument of dissipativity.

\begin{lemma}\label{lem:ergodicity} Let Assumption \ref{as:regularity} hold and let $b > 0$ be given in Assumption \ref{as:regularity}\ref{as:regularity_diss}. Then, if $\delta > 1/(4b)$, the process $(X^\epl, Z^\epl)^\top$ solution of \eqref{eq:systemSDE} is geometrically ergodic, i.e., there exists $C, \lambda > 0$ such that for all measurable $f$ such that for some integer $q > 0$ 
	\begin{equation}
		f(x) \leq 1 + \abs{x}^q,
	\end{equation}
	it holds
	\begin{equation}
		\abs{\E f(X^\epl_t, Z^\epl_t) - \int_\R\int_\R f(x, z) \rho^\epl(x, z) \dd x \dd z} \leq C\left(1 + \norm{\begin{pmatrix} X^\epl_0 & Z^\epl_0 \end{pmatrix}^\top}^q \right)e^{-\lambda t},
	\end{equation}
	where $\E$ denotes expectation with respect to the Wiener measure and $\rho^\epl$ is the solution to the stationary Fokker--Planck equation
	\begin{equation}
	\label{eq:FPsystem}
	\sigma \partial^2_{xx} \rho^\epl(x,z) +  \partial_x \left( \left( \alpha \cdot V'(x) + \frac{1}{\epl} p' \left ( \frac{x}\epl \right ) \right ) \rho^\epl(x,z) \right) + \frac{1}{\delta} \partial_z \left((z - x) \rho^\epl(x,z)\right) = 0.
	\end{equation}
\end{lemma}

\begin{proof} Lemma \ref{lem:density} guarantees that the Fokker--Planck equation can be written directly from the system \eqref{eq:systemSDE}. For geometric ergodicity, let 
\begin{equation}
	 \mathcal S(x,z) \defeq \begin{pmatrix} -\sum_{i=1}^N V_i'(x) \\ \frac1\delta (x - z) \end{pmatrix} \cdot \begin{pmatrix} x \\ z \end{pmatrix} = -\sum_{i=1}^N V_i'(x)x+\frac1\delta(xz-z^2)
\end{equation}
Due to Assumption \ref{as:regularity}\ref{as:regularity_diss} and Young's inequality, we then have for all $\gamma > 0$
\begin{equation}
	\mathcal S(x,z) \leq a + \left(\frac{1}{2\gamma\delta} - b\right)x^2 + \frac1\delta\left(\frac{\gamma}{2} - 1\right)z^2.
\end{equation}
We choose $\gamma = \gamma^* \defeq 1 - b\delta + \sqrt{2-2b\delta+b^2\delta^2} > 0$ so that
\begin{equation}
	C(\gamma^*) \defeq -\frac{1}{2\gamma^*\delta} + b = -\frac1\delta\left(\frac{\gamma^*}{2} - 1\right),
\end{equation}
and we notice that $C(\gamma^*) > 0$ if $\delta > 1/(4b)$. In this case, we have
\begin{equation}
	\mathcal S(x, z) \leq a - C(\gamma^*) \norm{\begin{pmatrix} x & z \end{pmatrix}^\top}^2,
\end{equation}
and problem \eqref{eq:systemSDE} is dissipative. The result then follows from \cite[Theorem 4.4]{MSH02}.
\end{proof}
